---
title: "PCoA_diversity_plots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(phyloseq)
library(DT)
library(ggplot2)
library(data.table)
library(pander)
library(cowplot)
library(ggpubr)
library(patchwork)
library(Hmisc)
library(gridExtra)
library(rstatix)
library(patchwork)
sgColorPalette = c("#84CF04","#01B5BB","#E50E63","#6D7272","#8F389E",
                     "#DF8236","#036B6B","#F1BA2F","#9F832D","#94E804",
                     "#01D4DB","#FAC131","#B0B8B8","#F08C3A","#FF106E",
                     "#B948CC","#05B5B5","#CFAA3A")
```

## Loading phyloseqs
```{r}
output_results = "../results/"
output_data = "../results/"
ps_deseq <- readRDS(paste0(output_results, "Filtered/ps_deseq_friedfilt.rds"))
ps_css <- readRDS(paste0(output_results,"Filtered/ps_css_friedfilt.rds"))
ps_no_norm <- readRDS(paste0(output_results,"Filtered/ps_no_norm_friedfilt.rds"))
ps_not_norm_comp <- readRDS(paste0("../data/ps_no_norm_age_bf_filt_complete_family.rds"))


#Remove Breastfed and their families
breast_fed <- c("089_A","054_N", "158_N" )

b_fed<-unique(ps_not_norm_comp@sam_data$Family.group.ID[which(ps_not_norm_comp@sam_data$Host.Name %in% breast_fed)])

#remove the whole family
for (i in b_fed){
  ps_not_norm_comp<- prune_samples(ps_not_norm_comp@sam_data$Family.group.ID != i, ps_not_norm_comp)

}

#Remove Possible Contradictions and their families
possible_phen_contra <- c("020_A","131_A", "184_A" )

p_con<-unique(ps_not_norm_comp@sam_data$Family.group.ID[which(ps_not_norm_comp@sam_data$Host.Name %in% possible_phen_contra)])

#remove the whole family
for (i in p_con){
  ps_not_norm_comp<- prune_samples(ps_not_norm_comp@sam_data$Family.group.ID != i, ps_not_norm_comp)

}

#For plots
ps_deseq@sam_data$Phenotype <- gsub("A", "ASD", ps_deseq@sam_data$phenotype)
ps_deseq@sam_data$Phenotype <- gsub("N", "TD", ps_deseq@sam_data$Phenotype)
ps_deseq@sam_data$Phenotype <-as.factor(ps_deseq@sam_data$Phenotype )

```

# CCA and visualization
Compare resulting amplicon data between and within sample types by canonical correlation analysis, regression profiling, and visualization (e.g. non-metric multi-dimensional scaling [NMDS], principle coordinates of analysis, principle component analysis).

## Constrained by status A or N
```{r cca_ind}
plotting_consPcoA <- function(ps, norm, variables){
  #drop samples with NA values in contraints
  keep <- as.vector(apply(ps@sam_data[, variables], 1, function(x) return(!any(is.na(x)))))
  ps <- prune_samples(keep, ps)
  
  #Ease of labeling 
  colnames(ps@sam_data)[colnames(ps@sam_data) == variables] <- gsub("frequency...longitudinal.", "", variables)
  variables <- gsub("frequency...longitudinal.", "", variables)
  colnames(ps@sam_data)[colnames(ps@sam_data) == variables] <- gsub("frequency.", "", variables)
  variables <- gsub("frequency.", "", variables)
  
  #include only families that have all 6 timepoints, 3 for ASD and 3 for NT participants, for ease of visualization
  fam_6<-names(table(sample_data(ps)$Family.group.ID)[table(sample_data(ps)$Family.group.ID) == 6])
  ps_6fam<-prune_samples(sample_data(ps)$Family.group.ID %in% fam_6,ps )
  sample_data(ps_6fam)$Family.group.ID <- paste0('fam', as.character(sample_data(ps_6fam)$Family.group.ID))
  formula <- as.formula(paste("~", variables))
  if(length(variables) > 1){
    formula <- as.formula(paste("~", paste(variables, collapse = "+")))
  }

  ps_pcoa_ord <- ordinate(
    physeq = ps_6fam, 
    method = "CAP", 
    distance = "bray",
    formula = formula
    )
  p <- plot_ordination(
    physeq = ps_6fam, 
    ordination = ps_pcoa_ord, 
    color = variables, 
    axes = c(1,2),
    title= paste("Constrained PcoA",norm,"ordinated by \n", paste(variables, collapse = "\n"))) + 
    geom_point( size = 2) +
    theme_minimal()+
    theme(text = element_text(size =10), plot.title = element_text(size=10), legend.title = element_blank())
  if(variables == "phenotype"){
    p <- p + scale_color_manual(values=sgColorPalette)
  }
  
  #sum_pcoA_DesEq<-summary(ps_pcoa_ord)
  erie_bray_sum_pcoA <- phyloseq::distance(ps, method = "bray")
  sampledf <- data.frame(sample_data(ps))
  beta_di<-betadisper(erie_bray_sum_pcoA, ps@sam_data$Family.group.ID)
  to_return<-list()
  to_return[[1]]<-p
  to_return[[2]]<-beta_di
  return(to_return)
}

```

## Constrained by phenotype
```{r}
#With Deseq
DeSeq_distance <- plotting_consPcoA(ps_deseq, "Deseq", variables = "Phenotype")
# plot
DeSeq_distance[[1]]

#same with CSS 
CSS_distance <- plotting_consPcoA(ps_css, "CSS", variables = "phenotype")
# plot
CSS_distance[[1]]
```

## Constrained by family
```{r cca_family}

#With Deseq
DeSeq_distance2 <- plotting_consPcoA(ps_deseq, "Deseq", variables = "Family.group.ID")
# plot
DeSeq_distance2[[1]]

#same with CSS 
CSS_distance <- plotting_consPcoA(ps_css, "CSS", variables = "Family.group.ID")
# plot
CSS_distance[[1]]

#the distance in those plot?
#average_distance_to_median
#pdf(file=paste0(output_data, "Figures/Distance_DeSeq_CSS_", Sys.Date(), ".pdf"))
boxplot(DeSeq_distance2[[2]]$distances,CSS_distance[[2]]$distances, names=c("DeSeq", "CSS"), 
        xlab = "Type of Normalization", ylab = "Distance on Component 1 & 2", main ="Intragroup distance for each family")
#dev.off()

```

## Constrained by top permanova variables
```{r fig.width = 12, fig.height = 9}
permanova_res <- read.csv(paste0(output_data, "PERMANOVA_noLR.csv"))
#scores for social, learning, and anxiety
social <- c("Language.ability.and.use", "Conversation.ability", "Plays.imaginatively.when.alone", "Plays.imaginatively.with.others", "Plays.in.a.group.with.others", "Eye.contact.finding", "Picks.up.objects.to.show.to.others", "Imitation.behavior")

learning <- c("Understands.speech", "Childhood.behavioral.development.finding")

anxiety <- c("Repetitive.motion", "Sleep.pattern.finding", "Response.to.typical.sounds", "Self.injurious.behavior.finding")

permanova_res <- permanova_res[-which(permanova_res$Variable %in% c(anxiety, social, learning)), ] #First only include 95 samples because optional question
n_sample_min <- 350
variables <- as.character(permanova_res[permanova_res$TotalN > n_sample_min & permanova_res$pvalue_adj <= 0.05, ]$Variable)
plot_list <- list()
for(variable in variables){
  res <- plotting_consPcoA(ps_deseq, "Deseq", variables = variable)
  plot_list[[variable]] <- res[[1]]
}

plot_list[[length(plot_list) + 1]]<- DeSeq_distance[[1]]
```

```{r, fig.width = 8.5, fig.height = 11}

pdf(paste0(output_data, 'pcoa_top_permanova_variables_vertical.pdf'), width = 8.5, height = 11, pointsize = 6)
plot_grid(plotlist = plot_list[c(1:7, length(plot_list))], ncol = 2)
dev.off()
```

```{r fig.width = 12, fig.height = 9}
pdf(paste0(output_data, 'pcoa_all_permanova_variables_vertical.pdf'), width = 8.5, height = 11, pointsize = 6)
plot_grid(plotlist = plot_list[c(1:8)], ncol = 2)
plot_grid(plotlist = plot_list[c(9:16)], ncol = 2)
plot_grid(plotlist = plot_list[c(17:20)], ncol = 2)
#plot_grid(plotlist = plot_list[c(25:length(plot_list))], ncol = 2)
dev.off()
```

# Diversity
Characterize and assess the diversity of each sample, and evaluate the extent of dissimilarity between the cohorts

## Alpha diversity
```{r alpha_diversity, fig.height=9.5, fig.width=7.5}
ER <- estimate_richness(ps_not_norm_comp, measures=c("Observed", "Chao1", "Shannon"))
ER <- cbind(ER, sample_data(ps_not_norm_comp)[row.names(ER), c("phenotype", "Family.group.ID", "Within.study.sampling.date")])
ER <- data.table(ER, keep.rownames = TRUE)

ER$phenotype <- gsub("A", "ASD", ER$phenotype)
ER$phenotype <- gsub("N", "TD", ER$phenotype)

ER_long <- melt(ER, id.vars=c('rn', 'phenotype', "Family.group.ID", "Within.study.sampling.date"))

# plot
ggplot(data=ER_long[variable!='se.chao1'], aes(x=phenotype, y=value, fill=phenotype))+
  geom_boxplot(width=0.7, outlier.colour='white')+
  geom_jitter(size=1, position=position_jitter(width=0.1))+
  xlab('')+ylab('')+
  scale_fill_manual(values=sgColorPalette)+
  theme_minimal()+facet_wrap(~variable, scales='free')



# run t-test to check significance
ttest=NULL
for(alphad in c('Observed', 'Chao1', 'Shannon')){
  ASD<-ER_long[phenotype == "ASD",]
  TD <-ER_long[phenotype == "TD",]
  ASD <- ASD[order(ASD$Family.group.ID, ASD$Within.study.sampling.date, ASD$variable),]
  TD <- TD[order(TD$Family.group.ID, TD$Within.study.sampling.date, TD$variable),]
  ER_long_order <- rbind(TD, ASD)

  ttest_res=t.test(value ~ phenotype, data=ER_long_order[ER_long_order$variable==alphad], var.equal=TRUE, paired = TRUE)
  ttest=rbindlist(list(ttest, data.table(alpha_index=alphad, pvalue=ttest_res$p.value)))
}

pander(ttest)

wilcox=NULL
for(alphad in c('Observed', 'Chao1', 'Shannon')){
  ASD<-ER_long[phenotype == "ASD",]
  TD <-ER_long[phenotype == "TD",]
  ASD <- ASD[order(ASD$Family.group.ID, ASD$Within.study.sampling.date, ASD$variable),]
  TD <- TD[order(TD$Family.group.ID, TD$Within.study.sampling.date, TD$variable),]
  ER_long_order <- rbind(TD, ASD)
  wilcox_res=wilcox.test(value ~ phenotype, data=ER_long_order[ER_long_order$variable==alphad], var.equal=TRUE)
  wilcox=rbindlist(list(wilcox, data.table(alpha_index=alphad, pvalue=wilcox_res$p.value)))
}

pander(wilcox)
twowayanova_alphad <- list()
for(alphad in c('Observed', 'Chao1', 'Shannon')){
ER <- estimate_richness(ps_not_norm_comp, measures=c("Observed", "Chao1", "Shannon"))
ER <- cbind(ER, sample_data(ps_not_norm_comp)[row.names(ER), c("phenotype", "Family.group.ID", "Within.study.sampling.date", "Host.Name")])
ER <- data.table(ER, keep.rownames = TRUE)

ER$phenotype <- gsub("A", "ASD", ER$phenotype)
ER$phenotype <- gsub("N", "TD", ER$phenotype)

ER_long <- ER

ER_long$Within.study.sampling.date <- gsub("Timepoint 1", "Timepoint_one", ER_long$Within.study.sampling.date)  
ER_long$Within.study.sampling.date <- gsub("Timepoint 2", "Timepoint_two", ER_long$Within.study.sampling.date) 
ER_long$Within.study.sampling.date <- as.factor(gsub("Timepoint 3", "Timepoint_three", ER_long$Within.study.sampling.date)) 
ER_long$phenotype <- as.factor(ER_long$phenotype)
#with wrapper
#res.aov <- anova_test(
#data = ER_long, dv = alphad, wid = Host.Name,
#between = phenotype, within = Within.study.sampling.date
#)
#res<-get_anova_table(res.aov)

#without wrapper formula
  res.aov <-aov(as.formula(paste0(alphad,"~ phenotype * Within.study.sampling.date + Error(Host.Name/(Within.study.sampling.date))")), data = ER_long)
  res<-summary(res.aov)
  res$Type <- alphad
  twowayanova_alphad<- c(twowayanova_alphad, res)
}
#No signficance

#MARA
ER <- estimate_richness(ps_not_norm_comp, measures=c( "Chao1", "Shannon"))
ER <- cbind(ER, sample_data(ps_not_norm_comp)[row.names(ER), c("Mobile.Autism.Risk.Assessment.Score", "phenotype", "Family.group.ID", "Within.study.sampling.date", "Age..months.")])
ER <- data.table(ER, keep.rownames = TRUE)
ER<-ER[ER$phenotype == "A",]
#ER_long <- melt(ER, id.vars=c('rn', "Mobile.Autism.Risk.Assessment.Score", "Family.group.ID", #"Within.study.sampling.date"))

# plot

chao<-ggplot(ER, aes(x=Mobile.Autism.Risk.Assessment.Score, y=Chao1, color=Chao1)) + geom_point()+ theme(legend.title = element_text("Chao1 Diversity Index")) + labs(color = "Chao1 Diversity Index", x = "Mobile Autism Risk Assessment Score", y = "Chao1 Diversity Index")
shan<-ggplot(ER, aes(x=Mobile.Autism.Risk.Assessment.Score, y=Shannon, color=Shannon)) + geom_point()+ theme(legend.title = element_text("Shannon Diversity Index")) + labs(color = "Shannon Diversity Index", x = "Mobile Autism Risk Assessment Score", y = "Shannon Diversity Index")
spearman.test(ER$Mobile.Autism.Risk.Assessment.Score, ER$Chao1, p=1)
spearman.test(ER$Mobile.Autism.Risk.Assessment.Score, ER$Shannon, p=1)
grid.arrange(shan, chao, ncol = 1)
shan<-shan+stat_cor(method="spearman",label.x = -6.75, label.y = 4.55, cor.coef.name = "R^2")
chao<-chao+stat_cor(method="spearman",label.x = -6.75, label.y = 400, cor.coef.name = "R^2")

stacked_plots <- shan + chao + plot_layout(ncol = 1)

stacked_plots 

age<-ggplot(ER, aes(x=Mobile.Autism.Risk.Assessment.Score, y=Age..months., color=Age..months.)) + geom_point()+ theme(legend.title = element_text("MARA and Age (Months)")) + labs(color = "Age in Months", x = "Mobile Autism Risk Assessment Score", y = "Age (Months)")
age+stat_cor(method="spearman", cor.coef.name = "R^2")

spearman.test(ER$Mobile.Autism.Risk.Assessment.Score, ER$Age..months., p=1)
```

## Beta diversity
```{r beta_diversity}
#Let's do a PcoA #not much differences 
GP.ord <- ordinate(ps_deseq, "PCoA", "bray")
p2 = plot_ordination(ps_deseq, GP.ord, type="samples", color="phenotype") +
  geom_point( size = 1)+
  scale_color_manual(values=sgColorPalette)+
  theme_minimal()
p2
```

```{r plot_wo_na}
# function to plot PCoA without NA points
wo_na_pcoa <- function(ps, pvar, ord_res){
  # ord_res: ordinated object
  
  keepid=!is.na(get_variable(ps, pvar)) &
    get_variable(ps, pvar)!='NA' &
    get_variable(ps, pvar)!=''
  tmp_ps <- prune_samples(keepid, ps)
  
  # get subset counts and metadata together
  DF <- cbind(ord_res$vectors[row.names(sample_data(tmp_ps)), 1:2], sample_data(tmp_ps)[, pvar])
  setnames(DF, pvar, 'testvar')
  
  # get eigenvalues
  eig=(ord_res$values$Eigenvalues/sum(ord_res$values$Eigenvalues))[1:2]*100
  
  p <- ggplot(data=DF, aes(x=Axis.1, y=Axis.2, colour=testvar))+
    geom_point(size=2)+
    ggtitle(pvar)+
    xlab(paste0('Axis.1 [', format(eig[1], digits=3), '%]'))+
    ylab(paste0('Axis.2 [', format(eig[2], digits=3), '%]'))+
    theme_minimal()+
    theme(legend.title=element_blank(), legend.position="bottom")
    
  print(p)
}
ord <- ordinate(ps_css, method = "PCoA", distance = "bray")
wo_na_pcoa(ps_css, 'Mobile.Autism.Risk.Assessment.Score', ord)
```

### Probiotics
```{r pcoa_probiotics, fig.height=5}
wo_na_pcoa(ps_deseq, 'Probiotic..consumption.frequency.', GP.ord)
```

### Antibiotics
```{r pcoa_antibiotics, fig.height=5}
# Anti.infective
wo_na_pcoa(ps_deseq, 'Anti.infective', GP.ord)

# Minimum.time.since.antibiotics
sample_data(ps_deseq)$Minimum.time.since.antibiotics <- as.numeric(sample_data(ps_deseq)$Minimum.time.since.antibiotics)
wo_na_pcoa(ps_deseq, 'Minimum.time.since.antibiotics', GP.ord)

```

### All other passed R2 cutoff and significant
```{r pcoa_permanova_sig, fig.height=5}
permanova_pcut = 0.05
permanova_cut = 0.02
permanova_res <- read.csv(paste0(output_data, "PERMANOVA_noLR.csv"))
for(pvar in permanova_res[permanova_res$R2>permanova_cut & permanova_res$pvalue<permanova_pcut]$Variable[-which(permanova_res[permanova_res$R2>permanova_cut & permanova_res$pvalue<permanova_pcut]$Variable %in% c("High_Anxiety"))]){
  wo_na_pcoa(ps_deseq, pvar, GP.ord)
}

```

### Check to see breastfed outliers on plot

```{r pcoa_permanova_sig3, fig.height=5}
ps_all<-readRDS("../results/Filtered/ps_allsamps_deseq_norm_friedfilt.rds")

breast_fed <- c("089_A","054_N", "158_N" )
ps_all@sam_data$breastfed <- rep("FALSE", length(rownames(ps_all@sam_data)))
ps_all@sam_data$breastfed[which(ps_all@sam_data$Host.Name %in% breast_fed)] <- rep("TRUE", length(ps_all@sam_data$breastfed[which(ps_all@sam_data$Host.Name %in% breast_fed)]))


GP.ord <- ordinate(ps_all, "PCoA", "bray")
p2 = plot_ordination(ps_all, GP.ord, type="samples", color="breastfed") +
  geom_point( size = 1)+
  scale_color_manual(values=sgColorPalette)+
  theme_minimal()
p2

GP.ord_con <- ordinate(ps_all, "CAP", "bray", formula = ~breastfed)
p2_con = plot_ordination(ps_all, GP.ord_con, type="samples", color="breastfed") +
  geom_point( size = 1)+
  scale_color_manual(values=sgColorPalette)+
  theme_minimal()
p2_con
saveRDS(p2, "../pcoa_diversity_plots_files/figure-html/unconstrained_pcoa_breastfed.rds")
saveRDS(p2_con, "../pcoa_diversity_plots_files/figure-html/constrained_pcoa_breastfed.rds")


ps_deseq_seafood<-subset_samples(ps_deseq, Seafood..consumption.frequency. != "NA")
GP.ord_con <- ordinate(ps_deseq_seafood, "CAP", "bray", formula = ~Seafood..consumption.frequency.)
p2_con = plot_ordination(ps_deseq_seafood, GP.ord_con, type="samples", color="Seafood..consumption.frequency.") +
  ggtitle(label = "Seafood Consumption (Ordinated)")
  #geom_point( size = 1)+
  #scale_color_manual(values=sgColorPalette)+
  #theme_minimal()
p2_con

ps_deseq_season<-subset_samples(ps_deseq, Season != "NA")
GP.ord_con <- ordinate(ps_deseq_season, "CAP", "bray", formula = ~Season)
p2_con = plot_ordination(ps_deseq_season, GP.ord_con, type="samples", color="Season") +
  ggtitle(label = "Seasonality")
  #geom_point( size = 1)+
  #scale_color_manual(values=sgColorPalette)+
  #theme_minimal()
p2_con

ps_deseq_income<-subset_samples(ps_deseq, Annual.household.income_rank != "NA")
GP.ord_con <- ordinate(ps_deseq_income, "CAP", "bray", formula = ~Annual.household.income_rank)
p2_con = plot_ordination(ps_deseq_income, GP.ord_con, type="samples", color="Annual.household.income_rank") +
  ggtitle(label = "Annual.household.income_rank")
  #geom_point( size = 1)+
  #scale_color_manual(values=sgColorPalette)+
  #theme_minimal()
p2_con

ps_deseq_income<-subset_samples(ps_deseq, Number.of.pet.reptiles != "NA")
GP.ord_con <- ordinate(ps_deseq_income, "CAP", "bray", formula = ~Number.of.pet.reptiles)
p2_con = plot_ordination(ps_deseq_income, GP.ord_con, type="samples", color="Number.of.pet.reptiles") +
  ggtitle(label = "Number.of.pet.reptiles")
  #geom_point( size = 1)+
  #scale_color_manual(values=sgColorPalette)+
  #theme_minimal()
p2_con
```

## Normalization comparison

```{r}
#Looking at NORMALIZATION
plotting_Fam_consPcoA <- function(ps,title){
  fam_6<-names(table(sample_data(ps)$Family.group.ID)[table(sample_data(ps)$Family.group.ID) == 6])
  ps_6fam<-prune_samples(sample_data(ps)$Family.group.ID %in% fam_6,ps )
  sample_data(ps_6fam)$Family.group.ID <- paste0('fam', as.character(sample_data(ps_6fam)$Family.group.ID))
  ps_pcoa_ord <- ordinate(
    physeq = ps_6fam, 
    method = "CAP", 
    distance = "bray",
    formula = ~ Family.group.ID
    )
  p<-plot_ordination(
    physeq = ps_6fam, 
    ordination = ps_pcoa_ord, 
    color = "Family.group.ID", 
    axes = c(1,2),
    title= paste("Constrained PcoA",title,"ordinated by families with all timepoints")
    ) + 
    geom_point( size = 2) +
    theme_minimal()+
    theme(text = element_text(size =10), plot.title = element_text(size=10), legend.position='none')
  
    #sum_pcoA_DesEq<-summary(ps_pcoa_ord)
    erie_bray_sum_pcoA <- phyloseq::distance(ps, method = "bray")
    sampledf <- data.frame(sample_data(ps))
    beta_di<-betadisper(erie_bray_sum_pcoA, sampledf$Family.group.ID)
to_return<-list()
to_return[[1]]<-p
to_return[[2]]<-beta_di
  return(to_return)
  }

#With Deseq
DeSeq_distance<-plotting_Fam_consPcoA(ps_deseq, "Deseq")
# plot
#DeSeq_distance[[1]]

#same with CSS 
CSS_distance<-plotting_Fam_consPcoA(ps_css, "CSS")
# plot
#CSS_distance[[1]]

#the distance in those plot?
#average_distance_to_median
#pdf(file=paste0(output_data, "Figures/Distance_DeSeq_CSS_", Sys.Date(), ".pdf"))
boxplot(DeSeq_distance[[2]]$distances,CSS_distance[[2]]$distances, names=c("DeSeq", "CSS"), 
        xlab = "Type of Normalization", ylab = "Distance on Component 1 & 2", main ="Intragroup distance for each family")
```

```{r}
ps_deseq_income<-subset_samples(ps_deseq, Number.of.pet.reptiles != "NA")
GP.ord_con <- ordinate(ps_deseq_income, "CAP", "bray", formula = ~Number.of.pet.reptiles)
p2_con = plot_ordination(ps_deseq_income, GP.ord_con, type="samples", color="Number.of.pet.reptiles") +
  ggtitle(label = "Number.of.pet.reptiles")
  #geom_point( size = 1)+
  #scale_color_manual(values=sgColorPalette)+
  #theme_minimal()
p2_con



```



